Comparative RNA-Seq: Signal to Noise Documentation

### 2022-03-07 manuscript.Rmd
Update to manuscript.Rmd
The methods section has been mostly finished. The running of each
inidivual tool, and the results need to be added in.

A data folder is added, and the data needs to be moved over and the
paths changed.

---
title: 'Comparative RNA-Seq: Signal to Noise Documentation'
author: "T.J Nicholson & P.P Gardner"
date: "March 2022"
output:
  html_document:
    toc: yes
    toc_depth: 4
    toc_float: yes
    number_sections: yes
  html_notebook:
    toc: yes
    toc_depth: 4
  pdf_document:
    toc: yes
    toc_depth: '4'
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggtree)
library(treeio)
library(ape)
library(UpSetR)
library(svglite)
library(ROSE)
library(ggpubr)
source('~/bin/R/functions.R')

@@

```{r toc, echo=FALSE, include=F} 
render_toc("manuscript.Rmd")
```

@@

```{bash, eval=F}

@@

@@

@@

```{r random_forest_frequency_plots, }
load("~/bin/PhD/Chapter_4/chapter_4_files/predDat.Rda")
load("~/bin/PhD/Chapter_4/chapter_4_files/validatation2.Rda")
#select the desired columns from the predicted data and validatation data

@@

```{r transcription_levels_by_result}

```{r random_forest_statistics, eval=F}
#TODO this is not used
load("~/bin/PhD/Chapter_4/chapter_4_files/predDat.Rda")
load("~/bin/PhD/Chapter_4/chapter_4_files/validatation2.Rda")

@@

```{r statistics_plots, eval=F}
#TODO this is not used
load("~/bin/PhD/Chapter_4/chapter_4_files/predDat.Rda")
load("~/bin/PhD/Chapter_4/chapter_4_files/validatation2.Rda")

@@

load("~/bin/Phd/Chapter_4/chapter_4_files/predDat.Rda")
load("~/bin/Phd/Chapter_4/chapter_4_files/validatation2.Rda")
##function written using max_dist as column name so each variable needs to  be renamed to this before using cumulativeDistribution()

>>>>

---
title: 'Comparative RNA-Seq: Signal to Noise Documentation'
author: "T.J Nicholson & P.P Gardner"
date: "March 2022"
output:
  pdf_document:
    toc: yes
    toc_depth: '4'
  html_notebook:
    toc: yes
    toc_depth: 4
  html_document:
    toc: yes
    toc_depth: 4
    toc_float: yes
    number_sections: yes
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggtree)
library(treeio)
library(ape)
library(UpSetR)
library(svglite)
library(ROSE)
library(ggpubr)
library(rstatix)
options(dplyr.summarise.inform = FALSE)
source('~/bin/R/functions.R')

@@

@@

```{r redundancy_1_R, eval=F}
#read in the data
dat <- read.table("predicted_contig_pos.txt", sep = " ", fill = T, as.is = T)
dat <- dat %>% select(V1, V4)
#split out the columns of interest
dat <- dat %>% separate(col = V1, into = c("contig", "coordinates"), sep = "\\/", remove = F)
dat <- dat %>% separate(col = coordinates, into = c("start", "stop"), sep = "-", remove = F)
dat <- dat %>% dplyr::rename(srna = V4) %>% select(contig, srna, start, stop) %>% mutate(strand = "+")
colnames(dat) <- c("contig","srna", "start", "stop", "strand")

#rearrange the start and stop so th at that the stop > start
dat <- dat %>% mutate(tmpstart = ifelse(start < stop, start, stop),
                      tmpend = ifelse(start > stop, start, stop))

#Set up the data to be used in genomic ranges overlap search
query <- GRanges(seqnames = dat$contig,
                 ranges = IRanges(start = dat$tmpstart, end = dat$tmpend),
                  strand = dat$strand, query_name = dat$srna)
lookup1 <- data.frame(id1 = dat$srna, queryHits = c(1:length(dat$srna)))
lookup2 <- data.frame(id2 = dat$srna, subjectHits = c(1:length(dat$srna)))

#Run search for overlaps and merge with original input to obtain pairs of 
#overlaps
tmp <- GenomicRanges::findOverlaps(query, query, type = 'any')
tmp <- as.data.frame(tmp)
tmp <- tmp %>% left_join(lookup1) %>% left_join(lookup2)
tmp <- tmp %>% filter(id1 != id2)
smallDat <- tmp %>% select(id1, id2) %>% unique()

#Swap columns and combine so that every combination is present in both columns
s2 <- smallDat
s2$id1 <- smallDat$id2
s2$id2 <- smallDat$id1
smallDat <- smallDat %>% bind_rows(s2) %>% unique()

ids <- unique(dat$srna)

#Loop over the ids and run a check to see if there are any overlapping ids
#to speed this up check if an id has already been seen and skip it.
#Write a list of overlapping ids to a unqiue file
checked_ids <- c()
for(item in ids){
  if(item %in% checked_ids){
    next
  }
  current_ids <- groupOverlapItems(smallDat, item, current_ids = c())
  checked_ids <- c(checked_ids, current_ids)
    write.table(x = current_ids, 
                file = paste("combined_alignment_ids/", 
                             current_ids[1], 
                             "_combined_list.txt", sep = ""), 
                append = T, quote = F, row.names = F, col.names = F)
}

@@

```{bash, eval=F}
#move into the folder where the lists of overlapping ids are kept
cd combined_alignments_ids/
mkdir -p fasta
mkdir -p hmm
mkdir -p alignments

#loop through the files that contain overlapping ids
for file in *.txt;
do
max_num="0"
outname=`basename $file _combined_list.txt`
#Check if there is only one ID listed in a file. If so there is no merge
#required.
line_count=`wc -l $file | cut -d ' ' -f1`
if (( $line_count == 1 ));then
cp ../large_alignments/alignments/$outname.stk ./alignments
continue
fi

#loop through the file and extract the sequences from each of the stk files
#that correspond to the ids that are overlapping
> fasta/$outname.fa
while read line; 
do 
esl-reformat fasta ../large_alignments/alignments/$line.stk >> fasta/$outname.fa
current_num=`esl-alistat ../large_alignments/alignments/$line.stk | grep "Number of sequences" | cut -d ' ' -f 4`

#check if the current alignment is the longest
if(( $current_num > $max_num ));
then
max_seq="$line.stk"
max_num="$current_num"
fi
done < $file

#build an hmm using the longest alignment
hmmbuild hmm/$outname.hmm ../large_alignments/alignments/$max_seq

#build an alignment using all the sequences from the overlapping ids
hmmalign --informat fasta hmm/$outname.hmm fasta/$outname.fa | esl-alimask -g --gapthresh 0.8 -p --pfract 0.5 --pthresh 0.5 - | esl-alimanip   --lnfract 0.6 --lxfract 1.4 --lmin 50 --lmax 500 --detrunc 50 - > alignments/$outname.stk
done
```

### File setup

-   There are some files that are consistently used throughout the analysis
  -   `genome_contig_pairs.txt` is a file that maps the genome `GCA_` ids to the 
      contig `NC` ids.
    -   this is necessary due to the alignment reporting contig names, but the 
        predicted RUFs belonging to genomes rather than contigs.
  -   `query_target_pairs.txt` consists of each alignment ID and the contig IDs
      of each sequence within the alignment.

```{bash,  eval=F}
#loop through all the genomes in the analysis
cd representative_genomes/
> ../genome_contig_pairs.txt
for file in GC*.fna;  
do   
  ID=`echo $file | cut -d '.' -f1,2 | cut -d "_" -f1,2`;   
  grep ^">" $file | cut -d ' ' -f1 | sort | uniq | sed 's/>//g' | sed -e "s/$/   $ID/" >> ../genome_contig_pairs.txt;   
done
```

```{bash query_target_pairs, eval=F}
#for each group loop through all the alignments and extract the contig IDs
#of all the sequences within each alignment
group="positive_control"
cd ${group}/alignments
> ../query_target_pairs.txt
for file in *.stk; 
do 
  ID=`echo $file | cut -d '.' -f1,2 | cut -d "_" -f1,2`; 
  ID_2=`echo $file | cut -d '.' -f1,2 `;  
  grep ^"#=GS" $file | sort | uniq | cut -d "/" -f1 | cut -d ' ' -f2 | rev | cut -d '|' -f1 | rev | sed -e "s/$/   $ID   $ID_2/" >> ../query_target_pairs.txt;  
done
```

@@

-   The file was not formatted in a usable way for reading into R
      -   `cat  RF00177.dnadist | tr '\n' ' ' | sed 's/ N/\nN/g' > RF00177.dists`

```{r, eval=FALSE}
#open the dnadist file
dat <- read.table("RF00177.dists", sep = "",
                  header = F, fill = T, stringsAsFactors = F, as.is = T)
dat <- dat %>% filter(!is.na(V2))
colnames(dat)[1] <- "phylip_id"
colnames(dat) <- c("names", dat$phylip_id)

#reformat the data to use in subsequent steps
meltDat <- melt(dat)
colnames(meltDat) <- c("phylip_id", "query.name", "distance")

#open list of phylip ids (truncated) and contig ids
load(file = "current_ids.Rda")
current_ids <- current_ids %>% unique()

#open list of genome and contig ids
contig_labels <- read.table("genome_contig_pairs.txt")
colnames(contig_labels) <- c("id", "target.genome")

#combine the phylip ids with the genome ids
contig_labels <- contig_labels %>% 
  left_join(current_ids, by = "id") %>% 
  select(-b)

#combine the genome ids with the distances data for the first column
colnames(contig_labels) <- c("target.id", "target.genome", "phylip_id")
meltDat <- meltDat %>% left_join(contig_labels, by = 'phylip_id')

#combine the genome ids with the distances data for the second column
colnames(meltDat)[1:2] <- c("target.name", "phylip_id")
colnames(contig_labels) <- c("query.id", "query.genome", "phylip_id")
meltDat <- meltDat %>% left_join(contig_labels, by = 'phylip_id')
colnames(meltDat)[2] <- c("query.name")

#take the pairs of genomes and arrange based on distances (to account for
#multiple contigs). Sort these and select a single distance for each pair. 
meltDat <- meltDat %>%  group_by(target.genome, query.genome) %>% arrange(as.numeric(distance)) %>% mutate(row_num = row_number()) %>% filter(row_num == 1) %>% select(-row_num)
save(meltDat, file = "~/bin/R/r_files/distanceMelt.Rda")

#Reshape the data back to a matrix (not sure this was needed but helped for 
#visualising the problem)
mat <- reshape2::acast(data = meltDat %>% 
                         select(target.genome, query.genome, distance), 
                       formula = target.genome ~ query.genome)
distanceMat <- as.data.frame(mat)
save(distanceMat, file = "~/phd/RNASeq/distanceMat.Rda")

#include a save a load point as the previous steps were time consuming and
#this allows for troubleshooting the next steps
load("~/phd/RNASeq/distanceMat.Rda")

#get the maximum distances between pairs of sequences for each alignment.
max_dists_pred <- calculateMaximumDistances('predicted')
save(max_dists_pred, file = "~/bin/R/r_files/max_dists_pred.Rda")
max_dists_pc <- calculateMaximumDistances('pc')
save(max_dists_pc, file = "~/bin/R/r_files/max_dists_pc.Rda")
max_dists_nc <- calculateMaximumDistances('negative_control')
save(max_dists_nc, file = "~/bin/R/r_files/max_dists_nc.Rda")

@@

```{r random_forest_frequency_plots, }
load("~/bin/manuscript/data/predDat.Rda")
load("~/bin/PhD/Chapter_4/chapter_4_files/validatation2.Rda")
#select the desired columns from the predicted data and validatation data

@@

```{r transcription_levels_by_result}
load("~/bin/manuscript/data/predDat.Rda")

#create groups withing the predDat based on the probability thresholds
predDat <- predDat %>% 
  mutate(group = ifelse(probability <0.17, 
                        'transcriptional noise',
                        ifelse(probability <= 0.81, 
                               'possible RNA', 
                               'putative RNA')))
predDat$group = factor(predDat$group, levels = c("transcriptional noise",
                                                 "possible RNA", 
                                                 "putative RNA"))
#boxplot of transcription levels for each group
p <- ggplot(data = predDat) +
  geom_boxplot(aes(y = read.max.score, group = group, fill = group),
               outlier.alpha = 0.1) +
  scale_y_continuous(trans = 'log10')
p
if(FALSE){
  ggsave(filename = paste0(figure_path, "SVG/trascription-comparison.svg"),
         plot = p, width = 450, height = 307, units = "mm")
}
#run wilcox test comparing the transcriptional noise and the other groups
wilcox.test(x = predDat$read.max.score[predDat$group == "transcriptional noise"],
            y = predDat$read.max.score[predDat$group == "putative RNA"])
wilcox.test(x = predDat$read.max.score[predDat$group == "transcriptional noise"],
            y = predDat$read.max.score[predDat$group == "possible RNA"])
```

```{r random_forest_statistics, eval=F}
#TODO this is not used
load("~/bin/manuscript/data/predDat.Rda")
load("~/bin/PhD/Chapter_4/chapter_4_files/validatation2.Rda")

@@

```{r statistics_plots, eval=F}
#TODO this is not used
load("~/bin/manuscript/data/predDat.Rda")
load("~/bin/PhD/Chapter_4/chapter_4_files/validatation2.Rda")

@@

load("~/bin/manuscript/data/predDat.Rda")
load("~/bin/Phd/Chapter_4/chapter_4_files/validatation2.Rda")
##function written using max_dist as column name so each variable needs to  be renamed to this before using cumulativeDistribution()
